Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-07-21 paper.Rnw
Figures formatted nicely, lots of other fixes.
In particular, I went through all of the TODOs and patched them up,
except for those that will require more thinking/work/or input.
For instance, the montages are still not complete, but I elaborated
the todo's with figure references so that other readers won't be
weirded out.

Also, I fixed a bunch of references, and put in citet references
where necessary.

Lastly, in addition to many wording changes and spelling/grammar
fixes, I added in errorbars on the ADNI-Xval plot, changed plot
text and point sizes, and included p/t-values for Pearson
corr

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }

@@

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty2006,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data \citep{MallarChakravarty2012} is possible.  The
novelty of this manuscript is the extension our multi-atlas methodology to the
hippocampus using more than a single input atlas, while simultaneously limiting
the number of possible inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelling using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations.  Although Wolz et al. begin with an atlas library of 30 MR images,
this method could theoretically work using a much smaller atlas library.  In
their validation, LEAP was used to segment the whole hippocampus in the ADNI1
baseline dataset, achieving a mean Dice score of 0.85 with manual segmentations.

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate that the
applicability of their work using data from 10 subjects.  In the second method,
Yushkevich and colleagues \citep{Yushkevich2009} acquired and labelled
hippocampal subfields on high-resolution MRI data from post-mortem medial
temporal lobe samples and demonstrate accurate parcellations of hippocampal subfields in
other MRI-volumes using nonlinear registration guided using manually derived
hippocampus masks and specific landmarks. \todo{flow is terrible in this sentence...}

@@

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps.  First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propogated to the target image space. 
Third, the labels are combined into a single labelling by way of a label fusion method 
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in other
publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available. These labels have been generated using the
SNT tool from Medtronic Surgical Navigation Technologies, Louisville, CO (see
Supplementary Materials for detailed discussion of the manual segmentation
process used).

@@

Monte Carlo Cross-Validation (MCCV), also known as repeated random sub-sampling
cross-validation, consists of repeated rounds of validation conducted on a fixed
dataset \cite{Shao1993}. In each round, the dataset is randomly partitioned into
a training set and a validation set. The method to be validated is then given
the training data, and its output is compared with the validation set.

@@

A total of ten validation rounds are perfomed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
we explore are: atlas library size (1-9), template library size (1-20),
registration method (\ants or \animal), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote).  A total of $10 \times 69 \times 9 \times 20 \times 2
\times 3 = \num{7452000}$ validation rounds were conducted, resulting in a total
of $\num{1490400}$ segmentations analysed.

@@

The \animal algorithm carries out image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller FWHM. The final transformation is a set of local translations defined on
a bed of equally spaced nodes that were estimated through the optimization of
the correlation coefficient.  For the purposes of this work we used the
regularization parameters optimized in Robbins et al. \citep{Robbins2004},
displayed in table \ref{tab:ANIMAL-params}.

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as 
the gold standard to compare with.  The segmentation accuracy reported is averaged 
over the ten validation rounds for each parameter setting.

@@

\paragraph{SZ-FEP dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see Malla et
al. \citep{Malla2003}.

@@

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2).  All images from the SZ-FEP dataset are segmented by \mb.  The 
optimal size of template library, registration method, and label fusion method 
found in Experiment 1 are used.

\paragraph{Evaluation method}
The manual segmentation protocol used to segment the Winterburn atlases has some
neuroanatomical differences in comparison to the protocol used to segment the
SZFEP dataset. Therefore, rather than evaluate using an overlap metric, such as
DSC, which assumes a fixed segmentation protocol, we instead correlate whole
hippocampus volume between \mb and manual segmentations.

\todo{take description from below}

@@

\paragraph{Evaluation method}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for evaluation of by measuring overlap. 
That is, different segmentation protocols will necessarily produce segmentations
that do not perfectly overlap, and so the degree of overlap cannot be used soley 
to assess two segmentaiton methods that each use different protocols.

Instead, we assess the correlation in hippocampal volume between \mb and manual segmentation
(SNT). For reference, we also compute the correlation in hippocampal volume between
the existing, established automated segmentation methods: FSL, FreeSurfer, and 
MAPER.  Additionally, we evaluate the volume-related fixed and proportional 
biases in all segmentation methods using Bland-Altman plots \todo{Bland-Altman ref}.

\todo{references} with manual segmentations. 
\todo{"hippocampal volume" = mean hemispheric volume?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-COMPLETE
% Prepares the data in a the form needed for plotting.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<ADNI-volumes-prep,echo=F,cache=T>>=
means.complete   <- read.csv(gzfile('data/cache/ADNI1:means.complete.csv.gz'))
package_totals   <- read.csv(gzfile('data/cache/ADNI1:package_totals.csv.gz'))
@

@@

No marked difference in segmentation accuracy is seen when either \animal or \ants
registration is used with any number of atlases or templates.  In every parameter 
configuration, the use of \mb with \ants registration shows a pronounced increase
in segmentation accuracy over \mb with \animal registration.  Surprisingly, the 
label fusion method used does not significantly improve label accuracy, contrary 
to the findings of Aljabar et al. \cite{Aljabar2009} when using weighted voting
on much larger altas/template libraries. In the remainder of  this section, only 
results using the \ants registration algorithm and majority vote fusion will 
be shown.

@@

\todo{mention failures}

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy 
(Figure \ref{ADNI1-xval-variability}).  The size of template library 
necessary to reach a significant (p<0.5) decrease in variance and standard 
deviation grows with the size of atlas library used.  A template library of 19
images is sufficient to show significant decrease in variance and standard 
deviation for 3-7 atlases. 
\todo{Tris: only show variance OR standard deviation because otherwise the story gets complicated... why compare measures?}

It is interesting to note that with an even number of templates, \mb shows a
small decrease in performance relative to when one fewer template image is used.
See section \todo{ref} for a discussion of this behaviour. In the remainder of
this section, only results from odd-sized template libraries will be shown.

@@

\begin{figure}
<<ADN1-library-composition, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
# These composition datasheets describe the subjects and subject DX in the 
# template and atlas libraries for each voting run
atlas_comp = read.csv('data/a2a_ants/atlas_library_composition.csv')
templ_comp = read.csv('data/a2a_ants/template_library_composition.csv')
a2a.best = subset(all_data, atlases == 9 &
                    templates.mb == 19 &
                    reg_method=='ANTS' &
                    method.mb =='Majority Vote')

cumulative_dist = 
  ggplot(a2a.best, aes(x=k.mb)) + stat_ecdf() + scale_y_continuous(breaks=seq(0,1,0.1))

a2a.best = subset(a2a.best, k.mb > 0.75) # drop the guys that suck

a2a.best = merge(a2a.best, atlas_comp, by=c("timestamp","atlases"))
a2a.best = merge(a2a.best, templ_comp, by.x=c("timestamp","templates.mb"), by.y=c("timestamp","templates"))
a2a.best$templates_AD_percent = (a2a.best$templates_AD / 19) * 100
a2a.best$templates_MCI_percent = (a2a.best$templates_MCI / 19) * 100
a2a.best.melted = melt(a2a.best, measure.vars=c("templates_AD_percent", "templates_MCI_percent"))

k_by_composition = 
  ggplot(a2a.best.melted, aes(x=value, y=k.mb, colour=DX)) + facet_grid(variable ~ DX) + 
  geom_boxplot(aes(group=as.factor(value), colour=DX)) +  geom_smooth(method="lm")

snt.unilateral.vols = read.csv('data/a2a_snt_volumes.csv')
a2a.best = merge(a2a.best, snt.unilateral.vols)

# bland-altman plots
bland_mb_overall = with(a2a.best, bland_altman_plot(volume.snt, volume.mb, DX) + 
  geom_smooth(method="lm") + 
  scale_y_continuous(breaks=seq(-500, 500, 50)))

bland_ma_overall = with(a2a.best, bland_altman_plot(volume.snt, volume.ma, DX) + 
  geom_smooth(method="lm") + 
  scale_y_continuous(breaks=seq(-500, 500, 50)))

# just the trials with a high percentage of MCI in the template library
a2a.best.highMCI = subset(a2a.best, templates_MCI_percent > 50)

bland_mb_highMCI = with(a2a.best.highMCI, bland_altman_plot(volume.snt, volume.ma, DX) + 
  geom_smooth(method="lm") + 
  scale_y_continuous(breaks=seq(-500, 500, 50)))

@
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<ADN1-xval-k-mean, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ggplot(subset(all_data_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,geom='point',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \label{fig:ADNI1-xval-k_mean}
  \caption{
  {\bf Mean DSC of \mb segmentations across 69 ADNI1 subjects, by atlas and
  template library size, registration algorithm, and label fusion method.} 
  }
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}

<<ADN1-xval-k-mean-mb.ma, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ss = subset(all_data_mean, reg_method == 'ANTS' & method.mb == 'Majority Vote')
ss = subset(ss, (templates.mb*atlases) %% 2 == 1)
ma_lines = ddply(ss, c("atlases"), function (df) {data.frame(yint = mean(df$k.ma))})

ggplot(ss,
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,geom='point',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  geom_hline(yintercept=ma_lines$yint, colour = c("red", "orange", "green", "cyan", "purple"), linetype="dashed") + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \label{fig:ADNI1-xval-k_mean}
  \caption{
  {\bf Mean DSC of \mb segmentations across 69 ADNI1 subjects, by atlas and
  template library size, registration algorithm, and label fusion method.} 
  }
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: "increase" in mean kappa over multi-atlas
\begin{figure}
<<ADN1-Xval-k-diff,cache=TRUE,dependson='ADNI1-Xval-prep',fig.width=6,fig.height=4>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & 
                method.mb=="Majority Vote" & (templates.mb*atlases) %% 2 == 1), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide = "none") + 
  scale_linetype_discrete(guide = "none") + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference in mean DSC" ) +
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{The difference in mean DSC between \mb and multi-atlas segmentations for 
  a range of parameter settings.}
  \label{ADNI1-xval-k-diff}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this experiment \mb is applied to a dataset of images of first episode
schizophrenia parients, using the Winterburn atlases and a template library of
21 subject images selected at random.  Expert manual whole hippocampal
segmentations are used as gold standards.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }

@@

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty2006,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension our multi-atlas methodology to
the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelling using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations.  Although \citet{Wolz2010} begin with an atlas library of 30 MR
images, this method could theoretically work using a much smaller atlas library.
In their validation, LEAP was used to segment the whole hippocampus in the ADNI1
baseline dataset, achieving a mean Dice score of 0.85 with manual segmentations.

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate that the
applicability of their work using data from 10 subjects.  In the second method,
\citet{Yushkevich2009} acquired and labelled hippocampal subfields on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and specific
landmarks, they demonstrate accurate parcellation of hippocampal subfields in
unlabelled MRI volumes.

@@

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps.  First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single labelling by way of a label fusion method 
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in other
publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available \citep{Hsu2002}. These labels have been
generated using the SNT tool from Medtronic Surgical Navigation Technologies,
Louisville, CO (see Supplementary Materials for detailed discussion of the
manual segmentation process used).

@@

Monte Carlo Cross-Validation (MCCV), also known as repeated random sub-sampling
cross-validation, consists of repeated rounds of validation conducted on a fixed
dataset \citep{Shao1993}. In each round, the dataset is randomly partitioned into
a training set and a validation set. The method to be validated is then given
the training data, and its output is compared with the validation set.

@@

A total of ten validation rounds are performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
we explore are: atlas library size (1-9), template library size (1-20),
registration method (\ants or \animal), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote).  A total of $10 \times 69 \times 9 \times 20 \times 2
\times 3 = \num{7452000}$ validation rounds were conducted, resulting in a total
of $\num{1490400}$ segmentations analysed.

@@

The \animal algorithm carries out image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller FWHM. The final transformation is a set of local translations defined on
a bed of equally spaced nodes that were estimated through the optimization of
the correlation coefficient.  For the purposes of this work we used the
regularization parameters optimized in \citet{Robbins2004}, displayed in table
\ref{tab:ANIMAL-params}.

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as
the gold standard to compare with \citep{Hsu2002}.  The segmentation accuracy
reported is averaged over the ten validation rounds for each parameter setting.

@@

\paragraph{\FEPdataset dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see
\citet{Malla2003}.

@@

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2).  All images from the \FEPdataset dataset are segmented by \mb.  The 
optimal size of template library, registration method, and label fusion method 
found in Experiment 1 are used.

\paragraph{Evaluation method}
The manual and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated.  This poses a problem for
evaluation by measuring overlap.  That is, since different protocols will
necessarily produce segmentations that do not perfectly overlap, the degree of
overlap cannot be solely used to compare segmentation methods using different
protocols. In place of an overlap metric, we can assess the degree of correlation in
average hippocampal volume of the subjects produced by each method.
Specifically, Pearson correlation is used.

@@

\paragraph{Evaluation method}
As in Experiment 3, the manual (SNT) and Winterburn hippocampal segmentation
protocols differ in the neuroanatomical features delineated, and so we must
assess \mb by the degree of correlation of average hippocampal volume across all
subjects produced by \mb and by manual segmentation.  Specifically, Pearson
correlation is used.  For comparison, we also compute the correlation in
hippocampal volume between the existing, established automated segmentation
methods: FSL, FreeSurfer, and MAPER.  Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-COMPLETE
% Prepares the data in the form needed for plotting.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<ADNI-volumes-prep,echo=F,cache=T>>=
means.complete   <- read.csv(gzfile('data/cache/ADNI1:means.complete.csv.gz'))
package_totals   <- read.csv(gzfile('data/cache/ADNI1:package_totals.csv.gz'))
@

@@

No marked difference in segmentation accuracy is seen when either \animal or \ants
registration is used with any number of atlases or templates.  In every parameter 
configuration, the use of \mb with \ants registration shows a pronounced increase
in segmentation accuracy over \mb with \animal registration.  Surprisingly, the 
label fusion method used does not significantly improve label accuracy, contrary 
to the findings of \citet{Aljabar2009} when using weighted voting on much larger
atlas/template libraries. In the remainder of  this section, only results using
the \ants registration algorithm and majority vote fusion will be shown.

@@

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy 
(Figure \ref{ADNI1-xval-variability}).  The size of template library 
necessary to reach a significant ($p < 0.05$) decrease in variance and standard 
deviation grows with the size of atlas library used.  A template library of 19
images is sufficient to show significant decrease in variance and standard 
deviation for 3-7 atlases.

We have omitted results obtained when using an even number of atlases or
templates since with this configuration we found significantly decreased
performance. We believe this is as a result of an inherent bias in the majority
vote fusion method used (see Discussion).

@@

% \begin{figure}
% <<ADN1-library-composition, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
% # These composition datasheets describe the subjects and subject DX in the 
% # template and atlas libraries for each voting run
% atlas_comp = read.csv('data/a2a_ants/atlas_library_composition.csv')
% templ_comp = read.csv('data/a2a_ants/template_library_composition.csv')
% a2a.best = subset(all_data, atlases == 9 &
%                     templates.mb == 19 &
%                     reg_method=='ANTS' &
%                     method.mb =='Majority Vote')
% 
% cumulative_dist = 
%   ggplot(a2a.best, aes(x=k.mb)) + stat_ecdf() + scale_y_continuous(breaks=seq(0,1,0.1))
% 
% a2a.best = subset(a2a.best, k.mb > 0.75) # drop the guys that suck
% 
% a2a.best = merge(a2a.best, atlas_comp, by=c("timestamp","atlases"))
% a2a.best = merge(a2a.best, templ_comp, by.x=c("timestamp","templates.mb"), by.y=c("timestamp","templates"))
% a2a.best$templates_AD_percent = (a2a.best$templates_AD / 19) * 100
% a2a.best$templates_MCI_percent = (a2a.best$templates_MCI / 19) * 100
% a2a.best.melted = melt(a2a.best, measure.vars=c("templates_AD_percent", "templates_MCI_percent"))
% 
% k_by_composition = 
%   ggplot(a2a.best.melted, aes(x=value, y=k.mb, colour=DX)) + facet_grid(variable ~ DX) + 
%   geom_boxplot(aes(group=as.factor(value), colour=DX)) +  geom_smooth(method="lm") 
% 
% snt.unilateral.vols = read.csv('data/a2a_snt_volumes.csv')
% a2a.best = merge(a2a.best, snt.unilateral.vols)
% 
% # bland-altman plots
% bland_mb_overall = with(a2a.best, bland_altman_plot(volume.snt, volume.mb, DX) + 
%   geom_smooth(method="lm") + 
%   scale_y_continuous(breaks=seq(-500, 500, 50)))
% 
% bland_ma_overall = with(a2a.best, bland_altman_plot(volume.snt, volume.ma, DX) + 
%   geom_smooth(method="lm") + 
%   scale_y_continuous(breaks=seq(-500, 500, 50)))
% 
% # just the trials with a high percentage of MCI in the template library
% a2a.best.highMCI = subset(a2a.best, templates_MCI_percent > 50)
% 
% bland_mb_highMCI = with(a2a.best.highMCI, bland_altman_plot(volume.snt, volume.ma, DX) + 
%   geom_smooth(method="lm") + 
%   scale_y_continuous(breaks=seq(-500, 500, 50)))
% 
%   
% @
% \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<ADN1-xval-k-mean, cache=TRUE, dependson='ADNI1-Xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(all_data_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \label{fig:ADNI1-xval-k_mean}
  \caption{Mean Dice's Similarity Coefficient of \mb segmentations with manual
  segmentations (SNT) for 69 ADNI1 subjects vs. atlas and template library size,
  registration algorithm, and label fusion method. Points above zero indicate an
  \mb parameter settings yielding a higher mean DSC score than multi-atlas
  segmentation using the same number of atlases.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: "increase" in mean kappa over multi-atlas
\begin{figure}
<<ADN1-Xval-k-diff,cache=TRUE,dependson='ADNI1-Xval-prep'>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & 
                method.mb=="Majority Vote" & (templates.mb*atlases) %% 2 == 1), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Number of Atlases") +
  scale_alpha_continuous(guide = "none") + 
  scale_linetype_discrete(guide = "none") + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference in mean DSC" ) +
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Difference in mean Dice's Similarity Coefficient of \mb and
  multi-atlas segmentations with manual (SNT) segmentations for a range of atlas
  and template library sizes.}
  \label{ADNI1-xval-k-diff}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this experiment \mb is applied to a dataset of images of first episode
schizophrenia patients, using the Winterburn atlases and a template library of
21 subject images selected at random.  Expert manual whole hippocampal
segmentations are used as gold standards.
